---
title: "Probability Mass Functions"
output:
  slidy_presentation:
    font_adjustment: 4
date: "2022-09-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Previously... (1/3)

**Discrete Random Variables and Probability Mass Functions**

**[PSDR] Definition 3.3**

A discrete random variable is a random variable that takes integer values. A discrete random variable is characterized by its **probability mass function (PMF)**. The PMF $p$ of a random variable $X$ is given by 
$$p(x) = P (X=x).$$

**[PSDR] Theorem 3.1**

Let $p$ be the probability mass function of $X$.

1. $p(x) \ge 0 \text{ for all } x$.
2. $\sum_x p(x) = 1$.

## Previously... (2/3)

**The Bernoulli Probability Mass Function**

Let $X$ be a Bernoulli random variable with PMF given as

$$P(X=x; p) = p^x (1-p)^{1-x}.$$

**[PSDR] Definition 3.15**

A Bernoulli trial is an experiment that can result in two outcomes, which we will denote as “success” and “failure.” The probability of a success will be denoted $p$, and the probability of failure is therefore $1−p$.

## Previously... (3/3)

**The Binomial Probability Mass Function**

**[PSDR] Definition 3.17**

A random variable $X$ is said to be a binomial random variable with parameters $n$ and $p$ if

$$P(X = x; n, p) = \binom{n}{x} p^x (1-p)^{(n-x)} \hspace{5px} \text{ for } x = 0,1,2,\cdots, n$$

where $\binom{n}{x} = \frac{n!}{x!(n-x)!}$. We will sometimes write $X \sim Binom(n,p)$.
      
**[PSDR] Theorem 3.3**

If $X$ counts the number of successes in $n$ independent and identically distributed (iid) Bernoulli trials, each with probability of success $p$, then $X \sim Binom(n,p)$.

## Example 1

The following example scenarios CAN be modeled using a Binomial PMF because the sampling method is independent and the there are only two outcomes.

- A player rolls a pair of fair dice 10 times. The number $X$ of 7s rolled is recorded.

- In a class of 30 students, 55% are female. The instructor randomly selects 4 students The number $X$ of females selected is recorded.

- Let $X$ be the certain number of credit card transactions that are fraudulent.

- Testing a newly developed drug on whether it is effective or not. Let $X$ be the number of patients who are cured by the drug.

- Let $X$ be the number of spam emails. We want to know the probability of getting a spam email or non-spam email in a given day.

## Example 2 (1/3)

*What are the chances that you encounter bots on Twitter?*

Twitter bots are complicated but let's simplify. Say it is known that 5\% of all Tweets are spam bots or tweets from a fake account.

If you encounter 40 tweets in a given day, we can compute the probability that a certain number of those tweets are bots using a Binomial PMF. Here, we assume that each tweet encounter is independent and identically distributed Bernoulli trial.

Let $X$ be a binomial random variable with probability of success $p = 0.05$ and number of independent trials $n = 40$.

## Example 2 (2/3)

* What is the chance that you encounter exactly 3 bot tweets?

\begin{align*}
P(X = 3) = & \binom{40}{3} (0.05)^3 (1-0.05)^{(40-3)} \\
         = & 0.1851 \\
\end{align*}

That is 18.51% chance of encountering three bot tweets out of 40 tweets.

You can use the `dbinom` function in R to compute the Binomial PMF.

```{r echo=TRUE, message=FALSE, warning=FALSE}
dbinom(3,40,0.05)
```

## Example 2 (3/3)

* What is the chance that you encounter 1, 2, or 3 bot tweets ?

\begin{align*}
P(X \le 3) = & \sum_{x = 0}^3 \binom{40}{x} (0.05)^x (1-0.05)^{(40-x)} \\
             &  - \binom{40}{0} (0.05)^0 (1-0.05)^{(40-0)} \\ 
             & = 0.7333
\end{align*}

The summation term above is what's known as the Cumulative Density Function (CDF).

That is 73.33% chance of encountering at most three bot tweets out of 40 tweets.

You can use the `pbinom` function in R to compute the Binomial CDF.

```{r echo=TRUE, message=FALSE, warning=FALSE}
pbinom(3,40,0.05) - dbinom(0,40,0.05)
```

## The Cumulative Distribution Function (CDF)

**[PSDR] Definition 4.4**

The cumulative distribution function (cdf) associated with $X$ is the function  

$$F(x)= P (X ≤ x).$$

For a discrete random variable $X$ with PMF $P(X = x)$, the the above is written as

$$F(x) = \sum_{n=-\infty}^x P(X = n).$$

## Bernoulli CDF

Let $X \sim Bernoulli(p)$ where $p$ is the probability of success.

- PMF: 

$$P(X=x; p) = p^x (1-p)^{1-x}.$$

- CDF:

\begin{align*}
F(x) = & P(X \le x; p) \\
     = & \sum_{n=0}^x p^x (1-p)^{1-x} \\
\end{align*}

\[
F(x) = \begin{cases}
          0 & \text{ if } x < 0  \\
          1-p & \text{ if } 0 \le x < 1 \\
          1 & \text{ if } x \ge 1
       \end{cases}
\]

## Binomial CDF

Let $X \sim Binomial(np)$ where $p$ is the probability of success and $n$ be the number of trials $n$.

- PMF:

$$P(X = x; n, p) = \binom{n}{x} p^x (1-p)^{(n-x)} \hspace{5px} \text{ for } x = 0,1,2,\cdots, n$$

Use `dbinom` in R.

- CDF:

\begin{align*}
F(x) = & P(X \le x; p) \\
     = & \sum_{k=0}^x \binom{n}{k} p^x (1-p)^{(n-k)} \\
\end{align*}

Use `pbinom` in R.

## Example 3 (1/3)

Now, let's ask a slightly different question from the questions in Examples 1 and 2. 

Consider a scenario where you look at multiple tweets. You don't know how many tweets you want to look at. What is the probability of observing a 1st bot on the 3rd tweet that you see?

Again, it is known that 5\% of all Tweets are spam bots or tweets from a fake account.

\begin{align*}
  P(\text{1 bot tweet until 1st tweet}) & = 0.05 = \frac{1}{20} \\
  P(\text{2 bot tweets until 1st tweet}) & = \frac{19}{20}\frac{1}{20} \\
  P(\text{3 bot tweets until 1st tweet}) & = \frac{19}{20}\frac{19}{20}\frac{1}{20} = \frac{19^2}{20^3}
\end{align*}

The answer is $P(\text{3 bot tweets until 1st tweet}) = 0.045$. There is a 4.50% probability of observing a bot tweet in the 3rd tweet that you see. In other words, the probability of observing failures in the 1st and 2nd tweet that you see, and a success in the 3rd tweet.

## Geometric Random Variable (2/3)

What is the probability of observing a 1st bot tweet on $(x+1)$-th tweet that you see?

\begin{align*}
  P(\text{1 bot tweet until 1st tweet}) & = 0.05 = \frac{1}{20} \\
  P(\text{2 bot tweets until 1st tweet}) & = \frac{19}{20}\frac{1}{20} \\
  P(\text{3 bot tweets until 1st tweet}) & = \frac{19}{20}\frac{19}{20}\frac{1}{20} = \frac{19^2}{20^3}
  \cdots & \\
  P(\text{x bot tweets until 1st tweet}) & = \left(\frac{19}{20}\right)^{x-1} \frac{1}{20}
\end{align*}

## Geometric Random Variable (3/3)

In this example, the Geometric PMF is given as

$$P(X = x) = \left(\frac{19}{20}\right)^{x-1} \left(\frac{1}{20}\right)$$

where $X$ is a Geometric random variable, $x$ is the number of failures because the $(x+1)$-th iteration is the first success with probability $p = \frac{1}{20}$.

The geometric PMF models the probability of the number of failures can you have before getting a 1st success. In this example, observing a bot tweet is a "success".

## The Geometric Probability Mass Function

**[PSDR] Definition 3.22**

A random variable $X$ is said to be a geometric random variable with parameter $p$ if

$$P(X = x) = (1-p)^{x-1} p, \hspace{20px} x = 1, 2, 3, \cdots$$

**[PSDR] Theorem 3.5**

Let $X$ be the random variable that counts the number of failures before the first success in a Bernoulli process with probability of success $p$. Then $X$ is a geometric random variable.

***Notice that the sample space is now infinite!***

## Geometric CDF

Let $X \sim Geometric(p)$ where $p$ is the probability of success.

- PMF: 

$$P(X = x; p) = (1-p)^{x-1} p, \hspace{20px} x = 1, 2, 3, \cdots$$

Use `dgeom` in R. Note that the input for `dgeom` is $(x-1)$. So if $x=3$, then you need `dgeom(3-1)`.

- CDF:

\begin{align*}
F(x) = & P(X \le x; p) \\
     = & \sum_{k=0}^x (1-p)^k p \\
\end{align*}

\[
F(x) = \begin{cases}
          1-(1-p)^{x} & \text{ if } x \ge 1  \\
          0 & \text{ if } x < 1 \\
       \end{cases}
\]

Use `pgeom` in R.

In your mini-assignment today, you are going to verify that the Geometric PMF is a valid PMF and the form of the Geometric CDF closed form is $F(x)$, shown above.

## Mini-Activity

[Mini-Assignment: Probability Mass Functions](ma-mth461a-20220922.pdf){target="_blank"}

Back to [Tentative Topics Schedule](../../topics.html).
